# Cayley Graphs {#lec3}
**Monday, January 25, 1993**

Given graphs $\Gamma = (X, E)$ and $\Gamma' = (X', E')$.

:::{.definition #isom}
A map $\sigma: X \to X'$ is an _isomorphism_\index{isomorophism} of graphs whenever;

i. $\sigma$ is one-to-one and onto,
ii. $xy\in E$ if and only if $\sigma x \sigma y \in E'$ for all $x, y\in X$.
:::

We do not distinguish between isomorphic graphs.

:::{.definition #auto}
Suppose $\Gamma = \Gamma'$. Above isomorphism $\sigma$ is called an _automorphism_\index{automorphism} of $\Gamma$. Then set $\mathrm{Aut}(\Gamma)$ of all automorphisms of $\Gamma$ becomes a finite group under composition.
:::

:::{.definition #transitive}
If $\mathrm{Aut}(\Gamma)$ acts transitive on $X$, $\Gamma$ is called _vertex transitive_\index{vertex transitive}.
:::


:::{.definition #cayley name="Cayley Graphs"}
Let $G$ be any finite group, and $\Delta$ any generating set for $G$ such that $1_G \not\in \Delta$ and $g\in \Delta \to g^{-1}\in \Delta$.
Then Cayley graph $\Gamma = \Gamma(G, \Delta)$ is defined on the vetex set $X = G$ with the edge set $E$ define by the following. \index{Cayley graph}

$$E = \{(h_1,h_2)\mid h_1, h_2\in G, h_1^{-1}h_2\in \Delta\} = \{(h, hg) \mid h\in G, g\in \Delta\}$$
:::

:::{.example #cyclic6}
$G = \langle a \mid a^6 = 1\rangle$, $\Delta = \{a, a^{-1}\}$.

```{r echo=FALSE, fig.width=2, fig.height=1.5, message=FALSE, fig.align="center"}
library(tidyverse)
df_eg31 <- data.frame(x = c(0, 0.5, 1.5, 2, 1.5, 0.5, 0), y = c(0, -1, -1, 0, 1, 1, 0))
ggplot(df_eg31, aes(x, y)) + geom_point() + geom_path() + theme_bw() + theme(panel.border = element_blank(),line = element_blank(),
        text = element_blank(),
        title = element_blank()) +
  annotate("text", x=-0.25, y = 0, label = "1") + annotate("text", x=0.33, y = -1, label = "a") +
  annotate("text", x=1.7, y = -0.95, label = "a^2", parse = TRUE) + annotate("text", x=2.2, y = 0, label = "a^3", parse = TRUE) +
  annotate("text", x=0.30, y = 1, label = "a^{-1}", parse = TRUE) + annotate("text", x=1.8, y = 0.98, label = "a^{-2}", parse = TRUE)
```
:::


:::{.example #k6minus1}
$G = \langle a \mid a^6 = 1\rangle$, $\Delta = \{a, a^{-1}, a^2, a^{-2}\}$.

```{r echo=FALSE, fig.width=2, fig.height=1.5, message=FALSE, fig.align="center"}
library(tidyverse)
df_eg32 <- data.frame(x = c(0, 0.5, 1.5, 2, 1.5, 0.5, 0.5, 2, 0.5, 0, 1.5, 1.5, 0), y = c(0, -1, -1, 0, 1, 1, -1, 0, 1, 0, -1, 1, 0))
ggplot(df_eg32, aes(x, y)) + geom_point() + geom_path() + theme_bw() + theme(panel.border = element_blank(),line = element_blank(),
        text = element_blank(),
        title = element_blank()) +
  annotate("text", x=-0.25, y = 0, label = "1") + annotate("text", x=0.33, y = -1, label = "a") +
  annotate("text", x=1.7, y = -0.95, label = "a^2", parse = TRUE) + annotate("text", x=2.2, y = 0, label = "a^3", parse = TRUE) +
  annotate("text", x=0.30, y = 1, label = "a^{-1}", parse = TRUE) + annotate("text", x=1.8, y = 0.98, label = "a^{-2}", parse = TRUE)
```
:::

:::{.example #d6}
$G = \langle a, b \mid a^6 = 1, b^2, ab = ba\rangle$, $\Delta = \{a, a^{-1}, b\}$.
```{r echo=FALSE, fig.width=3.5, fig.height=2.5, message=FALSE, fig.align="center", warning = FALSE}
library(tidyverse)
df_eg32 <- data.frame(x = c(0, 1, 3, 4, 3, 1, 0, 0.8, 1.5, 1, 1.5, 1.5, 2.5, 3, 2.5, 2.5, 3.2, 4, 3.2, 3.2, 2.5, 3, 2.5,2.5, 1.5, 1, 1.5, 1.5, 0.8), y = c(0, -1.5, -1.5, 0, 1.5, 1.5, 0, 0, -1, -1.5, -1, -1, -1, -1.5, -1, -1, 0, 0, 0, 0, 1, 1.5, 1, 1, 1, 1.5, 1, 1, 0))
ggplot(df_eg32, aes(x, y)) + geom_point() + geom_path() + theme_bw() + theme(panel.border = element_blank(),line = element_blank(),
        text = element_blank(),
        title = element_blank()) +
  annotate("text", x=-0.2, y = 0, label = "b") + annotate("text", x=1.6, y = -0.9, label = "a") +
  annotate("text", x=0.8, y = -1.5, label = "ab") + annotate("text", x=0.6, y = 1.56, label = "a^{-1}", parse = TRUE) +
  annotate("text", x=0.8, y = 1.53, label = "b") + annotate("text", x=3.2, y = -1.5, label = "a^2", parse = TRUE) +
  annotate("text", x=3.35, y = -1.53, label = "b") + annotate("text", x=3.2, y = 1.55, label = "a^{-2}", parse = TRUE) +
  annotate("text", x=3.39, y = 1.53, label = "b") +
  annotate("text", x=2.4, y = -0.85, label = "a^2", parse = TRUE) + annotate("text", x=1, y = 0, label = "1", parse = TRUE) +
  annotate("text", x=3, y = 0, label = "a^3", parse = TRUE) + annotate("text", x=4.2, y = 0, label = paste0("a^{3}"), parse=TRUE) +
  annotate("text", x=4.36, y = -0.02, label = paste0("b"), parse=TRUE) +
  annotate("text", x=1.6, y = 0.85, label = "a^{-1}", parse = TRUE) + annotate("text", x=2.4, y = 0.85, label = "a^{-2}", parse = TRUE)
```
:::


:::{.hs}
**HS MEMO**

$\mathrm{Aut}(\Gamma) \simeq D_6\times \mathbb{Z}_2$ contains two regular subgroups isomorphic to $D_6$ and $\mathbb{Z}_5 \times \mathbb{Z}_2$ and $\Gamma$ is obtained as Cayley graphs in two ways.
:::

Cayley graphs are vertex transitive, indeed.

:::{.theorem #cayley-graph}
The following hold.

| $(i)$ For any Cayley graph $\Gamma = \Gamma(G, \Delta)$, the map 
$$G \to \mathrm{Aut}(\Gamma) \; (g\mapsto \hat{g})$$
is an injective homomorphism of groups, where
$$\hat{g}(x) = gx \quad \textrm{for all }\: g\in G \textrm{ and for all } x\in X (= G).$$
Also, the image $\hat{G}$ is regular on $X$. i.e., the image $\hat{G}$ acts transitively on $X$ with trivial vertex stabilizers.

| $(ii)$ For any graph $\Gamma = (X, E)$, suppose there exists a subgroup $G \subseteq \mathrm{Aut}(\Gamma)$ that is regular on $X$. Pick $x\in X$, and let 
$$\Delta = \{g\in G\mid \langle x, g(x)\in E\}.$$
Then $1\not\in \Delta$, $g\in \Delta \to g^{-1}\in \Delta$, and $\Delta$ generates $G$. Moreover, $\Gamma \simeq \Gamma(G, \Delta)$.
:::

:::{.proof}
$(i)$ Let $g\in G$. We want to show that $\hat{g}\in \mathrm{Aut}(\Gamma)$. Let $h_1, h_2\in X = G$. Then,
\begin{align}
(h_1, h_2)\in E & \to h_1^{-1}h_2\in \Delta\\
  & \to (gh_1)^{-1}(gh_2)\in \Delta \\
  & \to (gh_1, gh_2)\in E\\
  & \to (\hat{g}(h_1), \hat{g}(h_2)) \in E.
\end{align}
Hence, $\hat{g}\in \mathrm{Aut}(\Gamma)$.

Observe: $g \mapsto \hat{g}$ is a homomorphism of groups:
$$\hat{1}_G = 1, \; \widehat{g_1g_2} = \widehat{g_1}\widehat{g_2}.$$

Observe: $g \mapsto \hat{g}$ is one-to-one:
$$\widehat{g_1} = \widehat{g_2} \to g_1 = \widehat{g_1}(1_G) = \widehat{g_2}(1_G) = g_2.$$

Observe: $\hat{G}$ is regular on $X$: Clear by construction.

| $(ii)$ $1_G\not\in \Delta$: Since $\Gamma$ has not loops, $(x, 1_Gx) \not\in E$.

$g\in \Delta \to g^{-1}\in \Delta$: 
$$g\in \Delta \to (x, g(x))\in E \to E \ni (g^{-1}(x), g^{-1}(g(x))) = (g^{-1}(x), x).$$

$\Delta$ generates $G$: Suppose $\langle \Delta \subsetneq G$. Let $\hat{X} = \{g(x)\mid g\in \langle \Delta\rangle\} \subsetneq X$. ($\hat{X} \subsetneq X$ as $G$ acts regularly on $X$.)

Since $\Gamma$ is connected, there exists $y\in \hat{X}$ and $z\in X\setminus \hat{X}$ with $yz\in E$. 

Let $y = g(x)$, $g\in \langle \Delta\rangle$, $z\in h(x)$, $h\in G\setminus \langle \Delta\rangle$. Then 
$$(y,z)=(g(x),h(x))\in E \to (x,g^{-1}h(x))\in E \to g^{-1}h\in \langle \Delta \rangle \to h\in \langle \Delta \rangle. $$
This is a contradition. Therefore, $\Delta$ generates $G$.

Let $\Gamma' = (X', E')$ denote $\Gamma(G, \Delta)$. We shall show that 
$$\theta: X' \to X \; (g\mapsto g(x))$$
is an isomorphism of graphs.

$\theta$ is one-to-one: For $h_1, h_2\in X' = G$,
$$\theta(h_1)=\theta(h_2) \to h_1(x) = h_2(x) \to h_2^{-1}h_1(x)=x \to h_2^{-1}h_1\in \mathrm{Stab}_G(x) = \{1_G\} \to h_1 = h_2.$$
($\mathrm{Stab}_G = \{g\in G\mid g(x) = x\}$.)

$\theta$ is onto: Since $G$ is transitive, 
$$X = \{g(x)\mid g\in G\} = \theta(X') = \theta(G).$$

$\theta$ respects adjacency: For $h_1, h_2\in X' = G$,
$$(h_1,h_2)\in E' \leftrightarrow h_1^{-1}h_2\in \Delta \leftrightarrow (x, h_1^{-1}h_2(x))\in E \leftrightarrow (h_1(x),h_2(x))\in E \leftrightarrow (\theta(h_1), \theta(h_2))\in E.$$
Therefore $\theta$ is an isomorphism between graphs $\Gamma(G, \Delta)$ and $\Gamma(X, E)$.
:::

How to compute the eigenvalues of the Cayley graph of and abelian group.

Let $G$ be any finite abelian group. Let $\mathbb{C}^*$ be the multiplicative group on $\mathbb{C}\setminus \{0\}$.

:::{.definition #character}
A (linear) $G$-character\index{character} is any group homomorphism $\theta: G \to \mathbb{C}^*$.
:::

:::{.example #cyclic3}
$G = \langle a\mid a^3 =1\rangle$ has three characters, $\theta_0, \theta_1, \theta_2$.
$$
\begin{array}{c|ccc}
\theta_i(a^j) & 1 & a & a^2 \\
\hline
\theta_0 & 1 & 1 & 1\\
\theta_1 & 1 & \omega & \omega^2\\
\theta_2 & 1 & \omega^2 & \omega
\end{array}, \quad \textrm{with }\; \omega = \frac{-1+\sqrt{-3}}{2}.
$$
Here $\omega$ is a primitive cube root pf $q$ in $\mathbb{C}^*$, i.e., $1+\omega + \omega^2 = 0$.
:::

For arbitraty group $G$, let $X(G)$ be the set of all characters of $G$.

Observe: For $\theta_1, \theta_2\in X(G)$, one can define product $\theta_1\theta_2:
$$\theta_1\theta_2(g) = \theta_1(g)\theta_2(g) \quad \textrm{for all }\; g\in G.$$
Then $\theta_1\theta_2\in X(G)$.

Observe: $X(G)$ with this product is an (abelian) group.

:::{.lemma #charactergroup}
The groups $G$ and $X(G)$ are isomorphic for all finite abelian groups $G$.
:::

:::{.proof}
$G$ is a direct sum of cyclic groups;
$$G = G_1\oplus G_2 \oplus \cdots \oplus G_m, \quad \textrm{where } \; G_i = \langle a_i\mid a_i^{d_i} = 1\rangle \quad (1\leq i\leq m).$$
Pick any alement $\omega_i$ of order $d_i$ in $\mathbb{C}^*$, i.e., a primitive $d_i$-the root of $1$. Define
$$\theta_i: G \to \mathbb{C}^* \quad (a_1^{\varepsilon_1}\cdots a_m^{\varepsilon_m} \mapsto \omega_i^{\varepsilon_i} \quad \textrm{where }\; 0\leq \varepsilon_i < d_i, 1\leq i\leq m).$$
Then $\theta_i\in X(G)$. (Exercise)

Claim: There exists an isomorphism of groups $G \to X(G)$ that sends $a_i$ to $\theta_i$.

Observe: $\theta_i^{d_i} = 1$. For every $g = a_1^{\varepsilon_1}\cdots a_m^{\varepsilon_m} \in G$,
$$\theta_i^{d_i}(g) = (\theta_i(g))^{d_i} = (\omega_i^{\varepsilon_i})^{d_i} = (\omega_i^{d_i})^{\varepsilon_i} = 1.$$
Observe: If $\theta_1^{\varepsilon_1}\theta_2^{\varepsilon_2}\cdots \theta_m^{\varepsilon_m} = 1$ for some $0\leq \varepsilon_i < d_i, 1\leq i\leq m$. Then $\varepsilon_1 = \varepsilon_2 = \cdots = \varepsilon_m = 0$.

_Pf._ $1 = \theta_1^{\varepsilon_1}\theta_2^{\varepsilon_2}\cdots \theta_m^{\varepsilon_m}(a_i) = \omega_i^{\varepsilon_i}$, Since $\omega_i$ is a primitive $d_i$-th root of $1$, $\varepsilon_i = 0$ for $1\leq i\leq m$.

Observe: $\theta_1, \ldots, \theta_m$ generate $X(G)$. Pick $\theta\in X(G)$. Since
$a_i^{d_i} = 1$, $1 = \theta(a_i^{d_i}) = \theta(a_i)^{d_i}$. 

Hence $\theta(a_i) = \omega^{\varepsilon_i}$ for some $\varepsilon_i$ with $0\leq \varepsilon_i < d_i$.

Now $\theta = \theta_1^{\varepsilon_1}\cdots \theta_m^{\varepsilon_m}$, since these are both equal to $\omega_i^{\varepsilon_i}$ at $a_i$ for $1\leq i \leq m$.

Therefore,
$$G \to X(G) \quad (a_i \mapsto \theta_i)$$
is an isomorphism of groups.
:::

**Note.** The correspondence above is clearly a group homomorphism.

