# Thin Modoles of a DRG {#lec16}
**Wednesday, February 24, 1993**

Let $\Gamma = (X, E)$ denote any graph of diameter $D$.

:::{.definition #ith-incidence}
For all integers $i$, the $i$-th incidence matrix $A_i\in \mathrm{Mat}_X(\mathbb{C})$ satisfies
$$(A_i)_{xy} = \begin{cases} 1 & \text{if $\partial(x,y) = i$,}\\
0 & \text{if $\partial(x,y)\neq i$,}
\end{cases} \quad (x,y\in X).$$
Observe, 
\begin{align}
A_0 & = I && (\text{identity})\\
A_1 & = A && (\text{adjacency matrix})\\
A_0 + A_1 + \cdots + A_D & = J && (\text{all 1's matrix}).
\end{align}
In general, $A_i$ may not belong to Bose-Mesner algebra.
:::


:::{.lemma #incidence-matrices}
Assume $\Gamma = (X, E)$ is distance-regular with diameter $D\geq 1$ and intersection numbers $c_i, a_i, b_i$.

| $(i)$ $$AA_i = c_{i+1}A_{i+1} + a_iA_i + b_{i-1}A_{i-1}, \quad (0\leq i\leq D, \; A_{-1} = A_{D+1} = O).$$

| $(ii)$ ${\displaystyle A_i = \frac{p_i(A)}{c_1c_2\cdots c_i}, \quad (0\leq i\leq D)}$, where $p_0, p_1, \ldots, p_D$ are polynomials for the trivial module from Lemma \@ref(lem:thin-module-structure).

| $(iii)$ $A_0, A_1, \ldots, A_D$ form a bais for Bose-Mesner algebra $M$.

| $(iv)$ For all distances $h, i, j \quad (0\leq, i, j, h\leq D)$, and for all vertices $x, y\in X$ with $\partial(x,y) = h$, the constant
$$p^h_{i,j} = |\{z\in X\mid \partial(x,z) = i, \; \partial(y,z) = j\}|$$
depends only on $h, i, j$ and not on $x, y$.

| $(v)$ ${\displaystyle E_0 = \frac{1}{|X|}J.}$
:::

:::{.proof}
| $(i)$ Pick $x\in X$. Apply each side to $\hat{x}$, we want to show that
$$AA_i \hat{x} = c_{i+1}A_{i+1}\hat{x} + a_iA_i\hat{x} + b_{i-1}A_{i-1}\hat{x}.$$
\begin{align}
\mathrm{LHS} & = A\left(\sum_{y\in X, \partial(x,y) = i}\hat{y}\right)\\
& = c_{i+1}\left(\sum_{z\in X, \partial(x,z) = i+1}\hat{z}\right) + a_i\left(\sum_{z\in X, \partial(x,z) = i}\hat{z}\right) + b_{i-1}\left(\sum_{z\in X, \partial(x,z) = i-1}\hat{z}\right)\\
& = \mathrm{RHS}.
\end{align}

| $(ii)$ Recall (Lemma \@ref(lem:thin-module-structure))
$$Ap_i(A) = p_{i+1}(A) + a_ip_i(A) + b_{i-1}c_ip_{i-1}(A) \quad (0\leq i\leq D).$$
Dividing by $c_1c_2\cdots c_i$, we have
$$A\frac{p_i(A)}{c_1c_2\cdots c_i} = c_{i+1}\frac{p_{i+1}(A)}{c_1c_2\cdots c_{i+1}} + a_i\frac{p_i(A)}{c_1c_2\cdots c_i} + b_{i-1}\frac{p_{i-1}(A)}{c_1c_2\cdots c_i}.$$
So, $A_i$, $p_i(A)/(c_1c_2\cdots c_i)$ satisfy the same recurrence.

Also boundary condition,
$$A_0 = p_0(A) = I.$$
Hence,
$$A_i = \frac{p_i(A)}{c_1c_2\cdots c_i}\quad (0\leq i\leq D).$$

| $(iii)$ Since $E_0, E_1, \ldots, E_D$ form a basis for $M$, $\dim M = D+1$.

Observe $A_0, A_1, \ldots, A_D\in M$ by $(ii)$,
$A_0, A_1, \ldots, A_D$ are linearly independent, since $p_0, p_1, \ldots, p_D$ are linearly independent. 

Thus, $A_0, A_1, \ldots, A_D$ form a basis for $M$.


| $(iv)$ $A_0, A_1, \ldots, A_D$ form a basis for an algebra $M$,
\begin{equation}
A_iA_j = \sum_{\ell = 0}^Dp^\ell_{ij}A_\ell \quad \text{for some }\; p^\ell_{ij}\in\mathbb{C}. (\#eq:plij)
\end{equation}
Fix $h \quad (0\leq h\leq D)$. Pick $x, y\in X$ with $\partial(x,y) = h$. 

Compute $x, y$ entry in \@ref(eq:plij), 
\begin{align}
(A_iA_j)_{xy} & = \sum_{z\in X}(A_i)_{xz}(A_j)_{zy}\\
& = \sum_{z\in X, \partial(x,z)=i, \partial(y,z)=j} 1\cdot 1\\
& = |\{z\in X\mid \partial(x,z)=i, \partial(y,z)=j\}|.
\end{align}
On the other hand, 
$$\left(\sum_{\ell=0}^D p^\ell_{ij} A_\ell\right)_{xy} = p^h_{ij}(A_h)_{xy} = p^h_{ij}.$$

| $(v)$ $\frac{1}{|X|}J$ is the orthogonal projection onto $\mathrm{Span}(\delta) = E_0V$. Hence,
$$\frac{1}{|X|} = E_0.$$
This proves the assertions.
:::

:::{.theorem #endpoint1}
Let $\Gamma = (X, E)$ be distance-regular with diameter $D\geq 2$ and intersection numbers $c_i, a_i, b_i$. Pick a vertex $x\in X$. Let $W$ be a thin irreducible $T(x)$-module with endpoint $r = 1$ and diameter $d$ ($d = D-2$ or $D-1$). Set $\gamma_0 = a_0(W) + 1$.

| $(i)$ The scalars 
\begin{equation}
\gamma_i := \frac{c_2c_3\cdots c_{i+1}b_2b_3\cdots b_{i+1}\gamma_0}{x_1(W)x_2(W)\cdots x_i(W)}   \quad (0\leq i\leq d)
(\#eq:gamma)
\end{equation}
$a_i(W), x_i(W)$ are algebraic integers in $\mathbb{Q}[\gamma_0]$. In particular, if $\gamma_0\in \mathbb{Q}$, then $\gamma_i$, $a_i(W)$ and $x_i(W)$ are integers for all $i$.

| $(ii)$ The numbers, $\gamma_i, a_i(W), x_i(W)$ can all be determined from $\gamma_0$ and the intersection numbers of $\Gamma$ in order 
$$x_1(W), \gamma_1, a_1(W), x_2(W), \gamma_2, a_2(W), \ldots $$
using $(i)$,
\begin{equation}
x_i(W) = c_ib_i + \gamma_{i-1}(a_i + c_i - c_{i+1} - a_{i-1}(W)) \quad (1\leq i\leq D-1), (\#eq:xi)
\end{equation}
and
\begin{equation}
a_i(W) = \gamma_i - \gamma_{i-1} + a_i + c_i - c_{i+1} \quad (1\leq i\leq D). (\#eq:ai)
\end{equation}

:::

**Note.**
$$p_i = p_1^W + \gamma_{i-1}p^W_{i-1} - c_i(p_{i-1}^W + \gamma_{i-2}p_{i-2}^W), \; (\gamma_{-1} = -\gamma_{-2} = 0, \; 0\leq i\leq d+1).$$

:::{.proof}
Set
$$\tilde{A}_i = A_0 + A_1 + \cdots + A_i \quad (0\leq i\leq D).$$

Claim 1. $A\tilde{A}_i = c_{i+1}\tilde{A}_{i+1}+(a_i-c_{i+1}+c_i)\tilde{A}_i + b_i\tilde{A}_{i-1} \quad (0\leq i\leq D-1).$

_Proof of Claim 1._
\begin{align}
\text{LHS} & = \sum_{j=0}^i AA_j\\
& = \sum_{j=0}^i (c_{j+1}A_{j+1} + a_jA_j + b_{j-1}A_{j-1})\\
& = \sum_{j=0}^{i-1}A_j(c_j+a_j+b_j) + A_i(c_i+a_i) + A_{i+1}c_{i+1}\\
& = k(A_0 + \cdots + A_{i-1}) + (a_i+c_i)A_i + c_{i+1}A_{i+1}.
\end{align}
\begin{align}
\text{RHS} & = c_{i+1}(A_0 + A_1 + \cdots + A_{i-1} + A_i + A_{i+1})\\
& \qquad + (a_i - c_{i+1}+c_i)(A_0 + A_1 + \cdots + A_{i-1} + A_i)\\
& \qquad + b_i(A_0 + A_1 + \cdots + A_{i-1})\\
& = k(A_0 + \cdots + A_{i-1}) + A_i(a_i + c_i) + A_{i+1}c_{i+1}.
\end{align}
This proves Claim 1.

Now pick $0\neq w \in E^*_1(x)W$ and let
$$w = \sum_{z\in X, \partial(x,z)=1}\alpha_z\hat{z}.$$
Pick $y$, where $\alpha_y\neq 0$.

For all $i$ $(0\leq i\leq D)$, define
\begin{align}
B_i & = \tilde{A}_i(\hat{x}- \hat{y})\\
& = \sum_{z\in X, \partial(x,z)\leq i}\hat{z} - \sum_{z\in X, \partial(y,z)\leq i} \hat{z}\\
& = \sum_{z\in X, \partial(x,z)=i, \partial(y,z)=i+1}\hat{z} - \sum_{z\in X, \partial(y,z)=i+1, \partial(y,z)=i} \hat{z}.
\end{align}
Note that $B_D = O$, $B_0 = \hat{x}-\hat{y}$, and 
$$\langle B_0, w_0\rangle = -\alpha_y \neq 0.$$

From Claim 1, 
$$AB_i = c_{i+1}B_{i+1}+(a_i-c_{i+1}+c_i)B_i + b_iB_{i-1} \; (0\leq i\leq D), \; B_{-1} = O.$$
Let $p_0^W, \ldots, p^W_d$ denote polynomials for $W$ from Lemma \@ref(lem:thin-module-structure). So,
$$w_i = p_i^W(A)w \in E^*_{1+i}(x)W, \quad (0\leq i\leq d).$$

Claim 2. $\langle w_i, B_j \rangle = 0$ if $j\not\in \{i, i+1\}$, $(0\leq i\leq d, 0\leq j\leq D)$.

_Proof of Claim 2._
$$w_i\in E^*_{1+i}(x)W, \quad B_j\in E^*_j(x)W + E^*_{j+1}(x)W.$$
```{r echo=FALSE, fig.width=5, fig.height=1.7, message=FALSE, fig.align="center"}
library(tidyverse)
df_eg161 <- data.frame(
  lx = c(2,0,10,8,9,8,6,10,2,4,6), ly = c(0,3,3,3,2,3,0,0,0,3,0), 
  rx = c(13,15,19,17,13,17,15,14,15,14,15), ry = c(3,3,3,0,0,0,3,2,3,2,3),
  dx = c(10,13,15,19,21,17,13,10,7,10,7), dy = c(3,3,3,3,0,0,0,0,0,0,0))
ggplot(df_eg161) + 
  geom_path(aes(lx, ly)) + geom_path(aes(rx, ry)) + geom_path(aes(dx, dy), linetype = "dotted") +
  theme_bw() + theme(panel.border = element_blank(),line = element_blank(),
        text = element_blank(),
        title = element_blank()) +
  annotate("text", x=0, y = 3.3, label = "B[0]", parse=TRUE) + annotate("text", x=4, y = 3.3, label = "B[1]", parse=TRUE)  +
  annotate("text", x=8, y = 3.3, label = "B[2]", parse=TRUE) + annotate("text", x=19, y = 3.3, label = "B[D-1]", parse=TRUE) +
  annotate("text", x=2, y = -0.3, label = "w[0]", parse=TRUE) + annotate("text", x=6, y = -0.3, label = "w[1]", parse=TRUE) +
  annotate("text", x=17, y = -0.3, label = "w[D-2]", parse=TRUE) 
```
Vertical lines indicate possible non-orthogonality.

Compute 
\begin{equation}
\langle Aw_i, B_j\rangle = \langle w_i, AB_j\rangle, \quad (0\leq i\leq D, \; 0\leq j\leq D-1).(\#eq:wibj)
\end{equation}
\begin{align}
\text{LHS} & = \langle w_{i+1},B_j\rangle + a_i(W)\langle w_i,B_j\rangle + x_i(W)\langle w_{i-1},B_j\rangle\\
\text{RHD}& = b_j\langle w_i, B_{j-1}\rangle + (a_j-c_{j+1}+c_j)\langle w_i, B_j\rangle + c_{j+1}\langle w_i, B_{j+1}\rangle.
\end{align}
Evaluate for $i = j-2, j-1, j, j+1$.

Set $i = j-2$.
```{r echo=FALSE, fig.width=3, fig.height=1.7, message=FALSE, fig.align="center"}
library(tidyverse)
df_eg162 <- data.frame(
  x = c(0,8,7,8,7,5,3,1,0,1,0,8), y = c(3,3,3,2,3,0,3,0,1,0,0,0))
ggplot(df_eg162) + 
  geom_path(aes(x, y)) +
  theme_bw() + theme(panel.border = element_blank(),line = element_blank(),
        text = element_blank(),
        title = element_blank()) +
  annotate("text", x=3, y = 3.3, label = "B[j-1]", parse=TRUE) + annotate("text", x=7, y = 3.3, label = "B[j]", parse=TRUE)  +
  annotate("text", x=1, y = -0.3, label = "w[j-2]", parse=TRUE) + annotate("text", x=5, y = -0.3, label = "w[j-1]", parse=TRUE)
```
Then \@ref(eq:wibj) becomes
$$\langle w_{j-1}, B_j\rangle = b_j\langle w_{j-2},B_{j-1}\rangle \quad (2\leq j\leq D-1).$$
By induction,
$$\langle w_{j-1}, B_j\rangle = b_2b_3\cdots b_j\langle w_{0},B_{1}\rangle \quad (1\leq j\leq D-1).$$
Define
$$\gamma_0 = \frac{\langle w_0, B_1\rangle}{\langle w_0, B_0\rangle}.$$
(We will show $\gamma_0 = 1+a_0(W)$.)

Then,
\begin{equation}
\langle w_{j-1},B_j\rangle = b_2b_3\cdots b_j\gamma_0\langle w_0, B_0\rangle. (\#eq:wj-1bj)
\end{equation}

Set $i = j+1$. 
Then \@ref(eq:wibj) becomes
$$x_{j+1}(W)\langle w_j, B_j\rangle = c_{j+1}\langle w_0, B_{j+1}\rangle \quad (0\leq j\leq d).$$
Hence,
\begin{equation}
\langle w_j, B_j\rangle = \frac{x_1(W)\cdots w_j(W)}{c_1c_2\cdots c_j}\langle w_0, B_0\rangle \quad (0\leq j\leq d). (\#eq:wjbj)
\end{equation}

Set $i = j-1$. 
Then \@ref(eq:wibj) becomes
$$
\langle w_j, B_j\rangle + a_{j-1}(W)\langle w_{j-1}, B_j\rangle
 = (a_j-c_{j+1}+c_j)\langle w_{j-1},B_j\rangle + b_j\langle w_{j-1},B_{j-1}\rangle.
$$
Evaluate this using \@ref(eq:wj-1bj) and \@ref(eq:wjbj). $(\langle w_0, B_0\rangle \neq 0)$. Then we have
<!-- \begin{split} -->
<!-- \langle w_j, B_j\rangle + a_{j-1}(W)\langle w_{j-1}, B_j\rangle\\ -->
<!-- & = (a_j-c_{j+1}+c_j)\langle w_{j-1},B_j\rangle + b_j\langle w_{j-1},B_{j-1}\rangle. -->
<!-- \end{split} -->
$$\frac{w_1(W)\cdots x_j(W)}{c_1\cdots c_j}+(a_{j-1}(W)-a_j+c_{j+1}-c_j)b_2\cdots b_j\gamma_0 = b_j\frac{x_1(W)\cdots x_{j-1}(W)}{c_1\cdots c_{j-1}},$$
$$\left(\gamma_i:=\frac{c_2c_3\cdots c_{i+1}b_2b_3\cdots b_{i+1}\gamma_0}{x_1(W)x_2(W)\cdots x_i(W)} \right).$$
$$\frac{x_j(W)}{c_j}  = b_j + \frac{c_1c_3\cdots c_{j-1}b_2b_3\cdots b_{j}\gamma_0}{x_1(W)x_2(W)\cdots x_{j-1}(W)}(a_j+c_j-c_{j+1}-a_{j-1}).$$
So,
$$x_j(W)  = c_jb_j + \gamma_{j-1}(a_j+c_j-c_{j+1}-a_{j-1}(W)).$$
This proves \@ref(eq:xi).

Set $i = j$.
Then \@ref(eq:wibj) becomes
$$
a_j(W)\langle w_j, B_j\rangle + x_{j}(W)\langle w_{j-1}, B_j\rangle
 = (a_j-c_{j+1}+c_j)\langle w_{j},B_j\rangle + c_{j+1}\langle w_{j},B_{j+1}\rangle.
$$
$$
(a_j(W) - (a_j-c_{j+1}+c_j))\frac{x_1(W)\cdots x_j(W)}{c_1\cdots c_j}
x_j(W)b_2\cdots b_j\gamma_0 - c_{j+1}b_2\cdots b_{j+1}\gamma_0 = 0.
$$
Thus,
$$a_j(W)-(a_j-c_{j+1}+c_{j}) + \frac{c_1\cdots c_jb_2\cdots b_j\gamma_0}{x_1(W)\cdots x_{j-1}(W)} - \frac{c_1\cdots c_jc_{j+1}b_2\cdots b_{j+1}\gamma_0}{x_1(W)\cdots x_j(W)} = 0,$$
or
$$a_j(W) = a_j + c_j - c_{j+1} - \gamma_{j-1} + \gamma_j.$$
This proves \@ref(eq:ai).

Also by setting $i = j = 0$, we have
\begin{align}
a_0(W)\langle w_0, B_0\rangle & = (a_0-c_1+c_0)\langle w_0, B_0\rangle + c_1\langle w_0, B_1\rangle\\
& = -\langle w_0, B_0\rangle + \gamma_0\langle w_0, B_0\rangle.
\end{align}
Hence, 
$$\gamma_0 = 1 + a_0(W).$$
Both $a_i(W)$ and $x_i(W)$ are algebraic integers, since they are eigenvalues of matrices with integer entries, namely,
$$E^*_{i+1}(x)AE^*_{i+1}(x) \; \text{ and }\; E^*_i(x)AE^*_{i+1}(x)AE^*_i(x).$$

Also $\gamma_0 = 1+a_0(W)$ is an algebraic integer, and $\gamma_i - \gamma_{i-1}$ is an algebraic integer by \@ref(eq:xi).

Hence, $\gamma_i$ is an algebraic integer by induction.

This completes the proof of Theorem \@ref(thm:endpoint1).
:::


:::{.example name="D=2"}
$$D = 2 \Leftrightarrow \text{strongly regular}.$$
Free parameters are $k, a_1, c_2$. 
Let $W$ be an irreducible module of endpoint $1$.
The matrix representation of $A|_W$ is 
$$\begin{pmatrix}
a_0(W) & x_1(W)\\ 1 & a_1(W)
\end{pmatrix}.$$
$a_0(W)$: free.
\begin{align}
x_1(W) & = c_1b_1 + (a_0(W) + 1)(a_1 + c_1 - c_2 - a_0(W))\\
& = k - a_1 - 1 + a_1a_0(W) + a_0(W) - c_2a_0(W) - a_0(W)^2\\
& \qquad + a_1 + a - c_2 - a_0(W)\\
& = a_1a_0(W) - c_2a_0(W) + k - c_2 - a_0(W)^2,\\
\gamma_1 & = 0,\\
a_1(W) & = -(a_0(W)+1) + a_1 + c_1 - c_2\\
& = -a_0(W) + a_1 - c_2.
\end{align}

Then the matrix has eigenvalues $\theta, \theta_1$.
There is one feasible condition: $a_0(W)$ is an algebraic integer.
:::


:::{.example name="D=3"}
Free parameters $c_2, c_3, k, a_1, a_2$. The matrix representation becomes
$$
A|_{W} = \begin{pmatrix} a_0(W) & x_1(W) & 0 \\ 1 & a_1(W) & x_2(W)\\ 0 & 1 & a_2(W) \end{pmatrix}.$$
Here, $a_0(W)$ is free $(=\gamma - 1)$
\begin{align}
x_1(W) & = k - 1 - a_1 + \gamma_0(a_1 + 1 - c_2 - a_0(W))\\
& = \gamma_0(a_1 - c_2 - a_0(W)) + k - a_1 + a_0(W).
\end{align}
Set
$$\gamma_1(W) = \frac{c_2b_2\gamma_0}{x_1(W)}.$$
\begin{align}
a_1(W) & = \gamma_1 - \gamma_0 + a_1 + 1 - c_2\\
x_2(W) & = \gamma_1(a_2 - c_3 - a_1(W)) + c_2(\gamma_0 + b_1 - a_2 + a_1(W))\\
a_2(W) & = -\gamma_1 + a_2 + c_2 - c_2.
\end{align}
The matrix has eigenvalues, $\theta, \theta_2, \theta_3$.

There are two feasibility conditions; $\gamma_0, \gamma_1$ are algebraic integers.

For arbitrary $D$, there are $D-1$ feasibility conditions;
$\gamma_0, \gamma_1, \ldots, \gamma_{D-1}$ are algebraic integers.
:::

:::{.lemma #aixigi}
With the notation of Theorem \@ref(thm:endpoint1), suppose
$$f_W = \frac{k-\lambda}{k} \quad (\text{so, }\; a_0(W) = -1).$$
Then, 
\begin{align}
a_i(W) & = a_i + c_i - c_{i+1} \quad (0\leq i\leq D-1)\\
x_i(W) & = b_ic_i \quad (1\leq i\leq D-1)\\
\gamma_i(W) & = 0 \quad (0\leq i\leq D-1).
\end{align}
:::

:::{.proof}
Since $\gamma_0 = a_0(W) + 1$, $\gamma_i = 0$.
:::
